\titleNameLong{}

### 2015-12-08 FinalProject/Tex/paper/paper.tex
finished dynamic scheduling subsection

\section{Dynamic Scheduling} \label{sec:scheduling}
\subsection{Coordinated warp scheduling (CAWA)}

\subsection{Variable Warp Size (VWS)}
This section left as an exercise to the reader

\subsection{Fine-grain Hardware Worklists (HWWL)}
This section left as an exercise to the reader

\subsection{Equalizer}
This section left as an exercise to the reader

>>>>

\section{Dynamic Scheduling} \label{sec:scheduling}
\subsection{Critical Warps}
GPUs manage to hide latency by swapping out stalled warps, but current
performance is not good enough. This latency hiding is limited by the following
characteristics of heterogeneous applications: imbalanced workloads in parallel
warps, diverging branch behavior, irregular memory access patterns, and shared
cache contention \cite{CoordinatedWarpScheduling}. These imbalances and
irregularities result in differences in execution time of parallel warps. The
slowest warp in a group is called the ``critical'' Since warps are blocked for
synchronization purposes until all of them are finished, this needlessly extends
the life of all the other warps and limits the performance of the application as
a whole. The authors of Criticality-aware warp acceleration (CAWA) demonstrate
that the average difference in execution time between the fastest and slowest
warp is 45\% and can reach as high as 70\%.

CAWA seeks to reduce the disparity among execution time in parallel warps using
three components:
\begin{description}
  \item[Dynamic warp criticality prediction] This predictor attempts to identify
  the critical warp that is used in the next two components. This is
  accomplished by using a per-warp criticality counter, which is based on the
  degree of instruction count disparity, a measure of how much diverging
  branches differ, as well as the number of stall cycles between two consecutive
  instructions. These two values are combined into a single criticality counter.
  \item[A greedy criticality-aware warp scheduler] This scheduler is built on
  top of the two existing warp scheduling policies, and gives more time to
  critical warps (based on the criticality counter described above). Ties are
  broken by the age of the warp, giving priority to older warps.
  \item[Criticality-aware cache prioritization] This scheme prioritizes cache
  accesses. It reserves a fixed number of L1 data cache ways for critical warps.
  The best number of ways to reserve was determined by experimental analysis and
  found to be 8 out of 16. The cache prioritization also uses a signature-based
  hit predictor to predict whether a given cache line belongs to the critical
  warp.
\end{description}

Using an oracle to identify the critical warp could yield a 21\% performance
improvement on average. The use of CAWA resulted in an average improvement of
9.2\% across all applications. Applications that are more sensitive to
scheduling policies and cache accesses saw a 23\% improvement in performance.
The accuracy of the critical warp identifier was also measured. It successfully
identified the critical warp as a slow warp 73\% of the time, where slow warp
here means it is in the slower half.

\subsection{Variable Warp Size}
While some of the past sections have discussed added new execution objects to
perform work, this technique pertains to splitting and re-merging existing
warps. There are two types of divergence in GPU execution: control flow
divergence and memory divergence. The former deals with complex control flow in
large warps resulting in performance loss, while the latter is related to how
well memory requests are coalesced into single warps. When warps are too small,
horizontal locality of memory accesses is lost; this effect is termed ``memory
convergence slip.''

In addition, workloads can be classified into three groups depending on how
their performance is affected by warp size. Divergent warps perform poorly as
warp size increases, due to control flow divergence as explained above. Warp-
size insensitive warps see little to no performance changes with warp size.
Convergent warps actually see better performance as warp size increases, due to
coalescing execution and memory accesses into a single warp. Increasing warp
size is better for the latter two types, since larger warps have better energy
efficiency.

Recent applications are increasingly divergent
\cite{VariableWarpSize}: raytracing, molecular dynamics simulations, game
physics simulations, and graph processing are all common applications for GPGPU
and all contain many divergent warps. To combat this, warp sizes can be adjusted
dynamically. A minimum of 4 is a good floor value, since graphics workloads
commonly process ``quads'' (threads in groups of 4) and there are diminishing
returns for warp sizes any smaller than this. This technique works by splitting
the traditional datapath into 8 slices, where each slice can fetch, decode, and
issue independently. Slices are statically assigned threads in a linear fashion
(0-3 to slice 0, 4-7 to slice 1, etc.), and threads cannot move between slices.

Allowing for variable warp sizes does not alter the register file and requires
no changes to memory. A warp ganging unit is added that fetches and decodes once
for all the warps in a gang. The gang issue scheduler enforces lock-step
execution of all the slices within a gang. Independent warps can be added to
fill holes in gangs, and gangs can be split and reformed. Fetching and issuing
gives priority to gangs to ensure the maximum level of lockstep.

The authors of this technique varied gang scheduling policies, reformation
policies, splitting policies, and size. They found that inelastic variable warp
sizing, in which gangs are split based on their control path and are never
merged back together, had the best performance. They measured a 35\% increase in
performance on a 32-wide machine, using 8-wide warps, at the cost of only 2.5\%
area.

\subsection{Hardware Worklists}
There are two common approaches to mapping graph algorithms to GPUs. The first
is topology-driven, in which the current thread index is used to determine on
which nodes to operate. This approach launches a new kernel for every iteration
and visits all nodes. Since launching kernels is expensive and some or many
nodes do not have any useful work to do in a given iteration, this results in
inefficiencies. The other approach is data-driven, where shared software
worklists (SWWL) are used to determine which nodes have useful work for a given
iteration. Only one kernel is launched at the start, and it is refilled with
work as necessary each iteration. This approach avoids all of the useless work
found in the topology-driven method, but can require a very large worklist,
which makes it prone to memory contention or access irregularity. Because of
this drawback, data-driven mapping is impractical unless aggressive software
optimizations are made.

There are four weaknesses that data-driven optimizations attempt to improve:
memory contention on pulls, contention on pushes, load balancing, and work
efficiency. Double-buffering is the simplest optimization and reduces memory
contention by partitioning the worklist push and pull sections. Work chunking
reduces push contention by grouping multiple pushes into one operation,
amortizing the cost. Work donating works by donating excess work to other
threads within the same block, improving load balance. Variable kernel
configuration adjusts the number of threads spawned by a kernel call based on
the number of active nodes in the worklist.

Even with such optimizations, the data-driven approach is not ideal due to
remaining memory contention on pushes and poor load balancing. Using fine-
grained hardware worklists (HWWL) instead of software worklists is one method of
dealing with such drawbacks
\cite{UsingFineGrainHardwareWorklists}. This works by exposing the HWWL to the
software through simple additions to the ISA. This way an application utilizing
a SWWL can be easily adapted to use a HWWL. The microarchitecture of the GPU is
modified to support a HWWL, distributing work to each lane in the GPU. To enable
better load balancing, empty banks pull from other banks instead of exiting.
There are two policies for when to redistribute work: interval-based and on-
demand. The former periodically redistributes work at a given sampling interval.
This has the advantage of better load balancing, but a higher cost. The
alternative, on- demand redistribution, only redistributes when a bank pulls or
pushes. This saves energy but has poorer load balancing than the interval-based
policy.

In addition, there are three possible ways to redistribute the work. The first
is threshold-based, where greedy (more full) banks donate to the needy (less
full) banks. The second scheme is local sorting-based, in which each HWWL bank
indicates how much work it has. A work redistribution unit (WRU) in each SM
sorts its own banks and attempts to redistribute within itself. The global
sorting-based scheme is similar to the second, but uses a single WRU for the
entire GPU. This last approach is unrealistic due to scalability issues.

The results of this technique are promising. When compared to the SWWL, the HWWL
had a 67\% reduction in memory stalls and between a 1.2-2.4x increase in
performance. The area overhead of the HWWL banks was 2.5\% of the register file
core, which is a small price to pay.

\subsection{Dynamic Frequency and Parallelism Adjustment}
The last technique in this section deals with dynamically monitoring resource
requirements and adjusting the SM frequency, memory frequency, and number of
threads executing. A majority of GPU applications are bottlenecked by either a
lack of compute resources, memory bandwidth, or a limited data cache, and so
executing the maximum number of threads and the highest frequencies is not
always beneficial to performance. If a given resource is bottle-necked, the
other resources become underutilized. For example, the NVIDIA Fermi architecture
has 48KB of data cache. When spread out over the maximum number of threads that
can execute at the same time, this leaves only 30 bytes for each.

If the characteristics of an application's workload were known before execution,
the GPU could be tuned to execute the optimal number of threads at the optimal
SM and DRAM frequencies. Unfortunately, this is often not possible: workloads
are heavily dependent on the given application, its input, and the specific GPU.
Equalizer \cite{EqualizerTuningOfResources} is a runtime system that runs on
individual SMs, determines the resource requirements of kernels, and adjusts the
number of executing threads and frequencies of the SM and memory.

Equalizer determines the overall resource requirements of a kernel based on the
status of the warps within the kernel. Warps are classified into one of four
groups: active warps, which are not paused; warps that cannot be executed
because they are waiting on operands from memory; warps ready to issue to the
memory pipeline but that the Load/Store queue cannot accept this cycle; and
warps ready to issue to the arithmetic pipeline, but cannot because the pipeline
is saturated. Based off these classifications, Equalizer determines whether to
adjust any of the three parameters once per ``epoch'' (end of each execution
window). The number of thread blocks is only changed if it is appropriate to do
so for three consecutive epochs. New voltage and frequency values are submitted
to the Frequency Manager every epoch and adjusted if necessary. This manager
uses one of three discrete settings, where the low and high settings represent a
15\% reduction or increase, respectively. This model assumes a linear change in
voltage with frequency.

Equalizer operates in one of two modes: energy efficient or high performance.
The authors found that there were insignificant area and energy overheads from
additional hardware needed to implement Equalizer. In performance mode, an
overall 22\% performance boost was measured, while also saving 6\% energy. The
efficient mode saved 15\% energy and resulted in a 5\% performance boost.
